---
title: "Investage fips codes which don't last the entire period"
output: html_document
---

# First step to load packages and file locations etc.
```{r include=FALSE}
rm(list=ls())
 
# 1a Declare root directory, folder locations and load essential stuff
project.folder <- paste0(print(here::here()),'/')
source(paste0(project.folder,'0_00_create_folder_structure.R'))
source(paste0(functions.folder,'script_initiate.R'))
```

# load original pre-1990 data
```{r}
dat.pre.1990 <- read.dta(paste0(original_population_data,'countyPopulationsnewyears.dta')) %>%
  dplyr::filter(year<1990 & year> 1981 & age!=99) %>% # county fips only reliable 1982 onwards
  select(year,fips,stateFips,countyFips,sex,age,pop)
```

# Load formatted cdc bridged-race data
```{r}
filename_in = paste0(population.formatted.folder,
                           'vintage_',vintage,'/Bridged-Race Population Estimates 1990-2020 ',start_year_cdc,' ',end_year_cdc,'.csv')
dat.1990.onwards = read_csv(filename_in)
```

# combine pre-1990 and cdc bridged-race (1990 onwards) data
```{r}
dat = bind_rows(dat.pre.1990, dat.1990.onwards) %>%
  arrange(year,fips,sex,age)
```

# Establish first and last year of population data for each county fips code, as well as completeness of records
```{r}
dat_year_range = dat %>% group_by(fips) %>% summarise(year_min=min(year),year_max=max(year),no_years=length(unique(year)))
```

# Figure out which county fips codes do not have the entire range of data
```{r}
dat_missing = dat_year_range %>%
  filter(year_min>start_year_cdc | year_max<end_year_cdc | no_years<max(no_years))
```

# Print incomplete counties
```{r}
print(dat_missing)
```

# Save output
```{r}
filename_out = paste0(population.incomplete.folder,
                      'vintage_',vintage,'/Incomplete Counties 1982 ',end_year_cdc,'.csv')
write_csv(dat_missing,filename_out)
```