---
title: "Infer monthly population (race)"
output: html_document
---

# First step to load packages and file locations etc.
```{r include=FALSE}
rm(list=ls())
 
# 1a Declare root directory, folder locations and load essential stuff
project.folder <- paste0(print(here::here()),'/')
source(paste0(project.folder,'0_00_create_folder_structure.R'))
source(paste0(functions.folder,'script_initiate.R'))
```

# Load population that has been fixed to have consistent counties
```{r}
dat_all= data.frame()
  for(year_selected in c(1990:end_year_cdc)){
    if(year_selected<1990){
      filename_in = paste0(population.county.fixed.folder,
                        'pre_1990/pop_fixed_counties_race_', year_selected,'.csv')  
    }
    if(year_selected>=1990){
      filename_in = paste0(population.county.fixed.folder,
                        'vintage_',vintage,'/pop_fixed_counties_race_', year_selected,'.csv')
    }
    dat_year=read_csv(filename_in)
    dat_all = data.table::rbindlist(list(dat_all,dat_year))
    rm(dat_year)
}
```

# Assume each population value is for the end of June/beginning of July (month = 6) and get rid of state and county individual columns (all info in fips)
```{r}
dat_all = dat_all %>% 
  mutate(month=6) %>%
  select(-stateFips,-countyFips)
```

# Create complete grid of population
```{r}
years = sort(unique(dat_all$year))
fips = sort(unique(dat_all$fips))
sexes = sort(unique(dat_all$sex))
races = sort(unique(dat_all$race))
months = c(1:12)

complete.grid = expand.grid(year=years,fips=fips,sex=sexes,race=races,month=months) %>%
  arrange(year,fips,sex,race,month)
  
dat_complete = left_join(complete.grid, dat_all) %>%
  arrange(fips,sex,race,year,month)
```

# Infer monthly populations
```{r}
dat_complete = dat_complete %>%
  group_by(sex,race,fips) %>%
  mutate(pop = round(na.approx(pop, na.rm=FALSE))) %>%
  fill(pop, .direction = "downup") 
```

# Save output by year in GitHub project
```{r}
for(year_selected in c(min(dat_complete$year):end_year_cdc)){
  dat_year = dat_complete %>% dplyr::filter(year==year_selected)
  if(year_selected<1990){
    filename_out = paste0(population.race.processed.folder,
                      'pre_1990/pop_monthly_race_groups_', year_selected,'.csv')  
  }
  if(year_selected>=1990){
    filename_out = paste0(population.race.processed.folder,
                      'vintage_',vintage,'/pop_monthly_race_groups_', year_selected,'.csv')
  }
  write_csv(dat_year,filename_out)
}
```

# Save output by year in local location
```{r}

new_population_data.race = paste0(new_population_data,'race_groups/')

for(year_selected in c(min(dat_complete$year):end_year_cdc)){
  dat_year = dat_complete %>% dplyr::filter(year==year_selected)
  if(year_selected<1990){
    filename_out = paste0(new_population_data.race,
                      'pre_1990/pop_monthly_5_year_age_groups_', year_selected,'.csv')  
  }
  if(year_selected>=1990){
    filename_out = paste0(new_population_data.race,
                      'vintage_',vintage,'/pop_monthly_5_year_age_groups_', year_selected,'.csv')
  }
  write_csv(dat_year,filename_out)
}
```